# Daycare Costs and Unexamined Assumptions {#daycare}

Keywords: Skepticism, Census data API

## Case Description

Data journalism is a welcome trend, but it has not always been accompanied by a rise in thoughtful analysis. I saw in a 2015 article from [Vox] and I was annoyed that the article relied on a map just to show percentages (why not a table?) and then began noticing other problems, such as the lack of a date for the data.

![Vox article on day care](https://cdn.vox-cdn.com/thumbor/xf2ee3OS9VrbZRYnGdImMWJImEc=/0x0:1200x630/920x0/filters:focal(0x0:1200x630)/cdn.vox-cdn.com/uploads/chorus_asset/file/3751728/childcaremap.png)

```{r setup3,  echo=FALSE, results= 'asis', warning=FALSE}
library(knitr)
library(kableExtra)
library(readr)
library(scales)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")
cost <- read.csv("assets/DayCare.csv")
names(cost[1]) <- 'Dollars'
kable(cost, format.args = list(big.mark = ","), caption = "2014 Annual Costs for Infant Daycare") %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

[Source] Child Care Aware® of America

The costs come from a [source] that gives separate estimates for an infant, a four-year old and a school child at day-care centers and the [Vox] article doesn’t say which age group it used to calculate the percentages of median income that costs represent. Worse, it uses median *household* income, when the source used *median state income* (*see* [source] at *fn* 49, *p.* 2.). (Unfortunately the [source] cites to a Census data table that no longer exists.)

## With the right metrics

Do the [Vox] estimates agree with both the costs reported by [Cost of Daycare] and the Census 2014 median household income, which is what [Vox] used?

```{r   mhi, echo = FALSE, warning=FALSE, message=FALSE, results= 'asis'}
library(acs)
#mhi <- acs.fetch(2014, span = 1, geo.make(state = "*"), table.number = ("B19013"))
mhi_acs <- acs.fetch(2014, span = 1, geo.make(state = "*"), table.number = ("B19013")) 
mhi_df <- as.data.frame(mhi_acs@estimate) %>% tibble::rownames_to_column()
colnames(mhi_df) <- c('State', 'Median_household_income') 
mhi_tbl <- mhi_df %>% filter(State != 'Puerto Rico')
cost_mhi <- cost %>% inner_join(mhi_tbl, by = 'State') 
pctdc <- cost_mhi %>% mutate(Percentage = Cost/Median_household_income) %>% mutate(Percentage = round(Percentage*100))
kable(pctdc, format.args = list(big.mark = ","), caption = "2014 Cost of Day Care for infants and median household income") %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

Sources: [source] and 2014 U.S. Census ACS Table B19013

Wherever [Vox] derived its data on median household income, the percentages conflict with the official Census data.

Repeating the table using median *family* income of families with children under 18 changes the picture even more.

```{r  mhi2, echo = FALSE, warning=FALSE, message=FALSE, results= 'asis'}
mfi <- acs.fetch(2014, span = 1, geo.make(state = "*"), table.number = ("B19126"))
mfi_acs <- acs.fetch(2014, span = 1, geo.make(state = "*"), table.number = ("B19126")) 
mfi_df <- as.data.frame(mfi_acs@estimate[,3]) %>% tibble::rownames_to_column()
colnames(mfi_df) <- c('State', 'Median_family_income') 
mfi_tbl <- mfi_df %>% filter(State != 'Puerto Rico')
cost_mfi <- cost %>% inner_join(mfi_tbl, by = 'State') 
pctdcf <- cost_mfi %>% mutate(Percentage = Cost/Median_family_income) %>% mutate(Percentage = round(Percentage*100))
kable(pctdcf, format.args = list(big.mark = ","), caption = "2014 Cost of Day Care for infants and median family income for families with children under 18") %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

Sources: [Cost of Daycare] and 2014 U.S. Census ACS Table B19126

The table below compares the [Vox] table with this analysis.

```{r  diff, echo = FALSE, warning=FALSE, message=FALSE, results= 'asis'}
pctdif <- pctdc %>% inner_join(pctdcf, by = 'State') %>% dplyr::select('State', 'Percentage.x', 'Percentage.y')
colnames(pctdif) <- c('State', 'Vox', 'This_paper') 
pctdif <- pctdif %>% mutate(Difference = Vox - This_paper)
kable(pctdif, format.args = list(big.mark = ","), caption = "Differences in estates of 2014 Cost of Day Care data for infants as a percentage of median income between Vox article and this paper") %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

## Takeaway
> Доверяй, но проверяй

Trust, but verify. This is true of any data, but especially data that you can trace trace to an authoritative source. And it is critical to think about the choice of a benchmark to evaluate one set of data against another. Here, the costs of day care came from a credible source and seem reasonable, based on anecdotal evidence. If this were more than an exercise, it would be important to seek out alternative estimates.

The data used to provide the cost of day care as a percentage of median household income proved a poor choice. Not *all* households have children. There is, however, data not only for *family households,* but even for *families with children under 18.* Relatively few households residing in retirement communities have day care expenses; it also turns out that families with children under 18 have a *higher* median income than the *all household* group.

There may be a way to determine median income of families with one or more children under 6, since the number of those families is collected or estimated (between the full Census years), which would provide an even better metric.

The data you have is not always the data you need.

Data sources: 

* [Cost of Daycare], Appendix I: 2014 Average Annual Cost of Full-Time Child Care by State, *pp.* 53-54 in pdf

* U.S. Census American Community Survey for 2014, using the **R** package by Ezra Haber Glenn (2018). acs: Download, Manipulate, and Present American Community Survey and Decennial Data from the US Census. R package version 2.1.3. https://CRAN.R-project.org/package=acs

[Vox]: https://www.vox.com/2015/6/3/8721331/child-care-costs
[Cost of Daycare]: http://usa.childcareaware.org/wp-content/uploads/2016/05/Parents-and-the-High-Cost-of-Child-Care-2015-FINAL.pdf
[source]: http://usa.childcareaware.org/wp-content/uploads/2016/05/Parents-and-the-High-Cost-of-Child-Care-2015-FINAL.pdf
