I'll use three 10% samples to check if the results are consistent. The idea is to see if the porportion of results are *independent.* In formal terms, that will be the null hypothosis. The test statistic is the chi-square test of independence. I'll use the conventional *p-value* of 0.05 as the cutoff; anything less than that means that I cannot reject the null hypothesis that the performance categories and the factors are independent. Here are the contingency tables:

```{r contab, results="asis", echo = FALSE, warning=FALSE}

cs <- as.tibble(dbGetQuery(con, "SELECT perf, kcluster from loans"))
s1 <- cs %>% sample_n(9000,replace=TRUE)
s2 <- cs %>% sample_n(9000,replace=TRUE)
s3 <- cs %>% sample_n(9000,replace=TRUE)
kable(table(s1), format.args = list(big.mark = ","), caption = "Sample 1 contingency table of performance (row) by cluster (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
kable(table(s2), format.args = list(big.mark = ","), caption = "Sample 2 contingency table of performance (row) by cluster (column)")  %>% kable_styling(bootstrap_options = "striped", full_width = F) 
kable(table(s3), format.args = list(big.mark = ","), caption = "Sample 3 contingency table of performance (row) by cluster (column)")  %>% kable_styling(bootstrap_options = "striped", full_width = F) 
``` 

and the chi-square tests are variable, depending on the particular sample.

```{r chisq, results="asis", echo = FALSE, warning=FALSE}

print(chisq.test(table(s1)))

print(chisq.test(table(s2)))

print("Sample 3 chi square test for independence")
print(chisq.test(table(s3)))
``` 

The chi square tests show that for each of the samples, the pricipal components are independent of performance.



### What do other categorical contingency tables tell us?

One of the available categorical variables is credit *grade* from the [credit disclosure].

```{r gradevperf, results="asis", echo = FALSE, warning=TRUE}
library(tidyverse)
library(DBI)
library(RMySQL)
drv <- dbDriver("MySQL")
con <- dbConnect(drv, username="root", password="", dbname ="dlf", host="localhost")
cs <- as.tibble(dbGetQuery(con, "SELECT grade,perf FROM loans"))
# WHERE grade = 'A+' OR grade = 'AP'"))
#(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of credit grade (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))

``` 

The warning message

    Warning message:
    In chisq.test(table(s1)) : Chi-squared approximation may be incorrect

can be easy to overlook, but it has an important message: Some of the grades are too few to occupy the *expected value* slots of the chi-squared distribution.


We can restrict the selection to only the A- and AP categories easily enough.

```{r gradperf, results="asis", echo = FALSE, warning=FALSE}
cs <- as_tibble(dbGetQuery(con, "SELECT grade, perf FROM loans WHERE grade = 'A+' OR grade = 'AP'"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of credit grades A+ and AP (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

With a *p-value* greater than 0.05, the A- and AP credit grades are not independent of performance. 

Let's try another categorical variabe, which transaction contained each loan, compared to that loan's performance. The "deal" is not independent of performance.

```{r loanbydeal, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT deal,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of deal (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))

``` 

The initial interest rates are independent of performance.

```{r loanbymargin, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT orate,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of interest rate (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

The first payment date, indicating the time of origination, is not independent

```{r fpd, results="oasis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT fpd,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of first payment date (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

The loan type is independent of performance.

```{r loanbyltype, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT ltype, perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of loan type (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

Occupancy type is independent of performance.

```{r loanbyotype, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT otype,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of occupancy type (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

Loan documentation type is independent of performance..

```{r loanbydtype, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT dtype,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of documentation type (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

Loan purpose (purchase, rate/term refinance/ cash-out refinance) is independent of performance.

```{r loanbypurp, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT purpose,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of loan purpose (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 
Property type is not independent of perforannce.

```{r loanbyptype, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT ptype,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of property type (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 

The metropolitan region of the property is independent of performance.

```{r loanbymetro, results="asis", echo = FALSE, warning=FALSE}
cs <- as.tibble(dbGetQuery(con, "SELECT metro,perf FROM loans"))
kable(table(cs), format.args = list(big.mark = ","), caption = "Contingency table of metropolitan area (row) by performance (column)") %>% kable_styling(bootstrap_options = "striped", full_width = F) 
chisq.test(table(cs))
``` 


### Correlation analysis

For for discrete (or categorical) values, we can turn to contingency tables, where one variable is in the rows and the other in the columns, so this is sometimes called an *r x c* table.

